Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-11-22 ms/ms.Rmd

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

>>>>

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

## Parallel-oblique flaking

When compared with other large finished bifaces in the region, the distinctive parallel oblique flake scar patterns on both faces of the elliptical bifaces for all, or virtually all, of their length is unique (Figure \ref{fig:illustration}). The absence of identified preforms for this artifact type makes technological and/or attribute-based discriminations related to manufacture a challenge; however, one potential preform was recovered from mortuary contexts at Belcher Mound (Figure \ref{fig:illustration}c). The projection of ritual and/or symbolic interpretation/s that articulate with this flake scar pattern is not possible at this time. Parallel oblique flake scar patterns are known to be characteristic of Guerrero arrow points found at missions in Texas \citep{RN9610}, several dart point types of the late Paleoindian period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and this distinctive flaking pattern has also been inferred from ceramic motifs in Peru \citep{RN7938}.

Parallel oblique flake scar patterns are not characteristic of any of the larger bifaces and bifacial knives in the American Southeast; thus, technological, cultural, or symbolic relationships between the elliptical bifaces and Guerrero arrow points or Angostura dart points are not suggested here. The technological choice to use parallel oblique flaking in finishing these elliptical bifaces most likely represents, at minimum, a stylistic choice by knappers limited to this specific implement; although, that may have been a functionally-relevant choice given their shape and size. In some respects, the decision to employ parallel oblique flaking is as distinctive as the presence---although not diagnostic \citep{RN6170}---of the flat and expanding flake scar pattern observed on Gahagan and Copena bifaces.

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustration}
\end{figure}

### 2022-11-22 ms/ms.tex

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

>>>>

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/map.png}
\caption{Location of Millsap Cache, Jowell Farm, and other sites mentioned in the text. Elliptical bifaces have been found in every major drainage basin in the ancestral Caddo area (white).}
\label{fig:map}
\end{figure}

\hypertarget{parallel-oblique-flaking}{%
\subsection{Parallel-oblique flaking}\label{parallel-oblique-flaking}}

When compared with other large finished bifaces in the region, the
distinctive parallel oblique flake scar patterns on both faces of the
elliptical bifaces for all, or virtually all, of their length is unique
(Figure \ref{fig:illustration}). The absence of identified preforms for
this artifact type makes technological and/or attribute-based
discriminations related to manufacture a challenge; however, one
potential preform was recovered from mortuary contexts at Belcher Mound
(Figure \ref{fig:illustration}c). The projection of ritual and/or
symbolic interpretation/s that articulate with this flake scar pattern
is not possible at this time. Parallel oblique flake scar patterns are
known to be characteristic of Guerrero arrow points found at missions in
Texas \citep{RN9610}, several dart point types of the late Paleoindian
period \citep{RN8584,RN7935,RN11182, RN9843,RN8143,RN10117,RN11181}, and
this distinctive flaking pattern has also been inferred from ceramic
motifs in Peru \citep{RN7938}.

Parallel oblique flake scar patterns are not characteristic of any of
the larger bifaces and bifacial knives in the American Southeast; thus,
technological, cultural, or symbolic relationships between the
elliptical bifaces and Guerrero arrow points or Angostura dart points
are not suggested here. The technological choice to use parallel oblique
flaking in finishing these elliptical bifaces most likely represents, at
minimum, a stylistic choice by knappers limited to this specific
implement; although, that may have been a functionally-relevant choice
given their shape and size. In some respects, the decision to employ
parallel oblique flaking is as distinctive as the presence---although
not diagnostic \citep{RN6170}---of the flat and expanding flake scar
pattern observed on Gahagan and Copena bifaces.

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting parallel oblique flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting parallel oblique removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustration}
\end{figure}
